A full approximation scheme multilevel method \\ for solving nonlinear variational inequalities

### 2024-07-25 talk-ismp/slides.tex
fiddles

@@

	\begin{itemize}
	\item[$\circ$] $\mathcal{V}=H^1(\Omega)$, $F : \mathcal{V} \to \mathcal{V}'$ continuous, $\ell\in \mathcal{V}'$
	\end{itemize}
\item discretization\footnote{discretization = finite element (FE) method, throughout this talk} gives algebraic system on fine mesh $\Omega^h$:
    $${\color{FireBrick} F^h(u^h) = \ell^h}$$
\item suppose $w^h$ is a not-yet-converged iterate:
    $$r^h=\ell^h - F^h(w^h), \qquad \|r^h\| > \text{TOL}$$
\end{itemize}
}
\only<2>{
\begin{itemize}
\item how can we improve $w^h$ \emph{without} globally linearizing $F^h$?

	\begin{itemize}
	\item are there alternatives to Newton's method?
	\end{itemize}
\item rewrite the equation as the nonlinear \emph{correction equation}
    $${\color{FireBrick} F^h(u^h) - F^h(w^h) = r^h}$$
\item \underline{for $F^h$ linear}, convert the correction eqn to the \emph{error equation}
    $$F^h(e^h) = r^h,$$
with approximate solution $\tilde e^h$, and update $w^h \leftarrow w^h+\tilde e^h$
\end{itemize}
}
\only<3>{
\begin{itemize}
\item there are fast \emph{smoothers}: inexpensive algorithms which improve $w^h$ by efficiently removing high-frequency error components
	\begin{itemize}
	\item sweeping through and solving nodewise problems (nonlinear Gauss-Seidel) is a smoother
	\end{itemize}
\item how about low-frequency error components?
\item \emph{idea}: use a coarser mesh $\Omega^H$ to estimate $u^h$ in the correction eqn
\end{itemize}

\vspace{5mm}
}
\only<4>{
\begin{itemize}
\item Brandt's (1977) \emph{full approximation scheme} (FAS) equation:
	$${\color{FireBrick} F^H(u^H) - F^H(\iR w^h) = R \, r^h(w^h)}$$

    \begin{itemize}
    \item[$\circ$] $\iR:\mathcal{V}^h \to \mathcal{V}^H$ is node-wise \emph{injection}
    \item[$\circ$] $R:(\mathcal{V}^h)' \to (\mathcal{V}^H)'$ is \emph{canonical restriction}
    \item[$\circ$] note: if $w^h=u^h$ exactly then $u^H = \iR w^h$ since $F^H$ injective
    \end{itemize}

\item let $\ell^H = F^H(\iR w^h) + R\, r^h(w^h)$
\item rewritten FAS equation:
    $${\color{FireBrick} F^H(u^H) = \ell^H}$$
\end{itemize}
}
\end{frame}

@@

\begin{center}
fine mesh $=\Omega^h$\, \includegraphics[height=0.14\textheight]{../talk-oxford/images/fine-grid.png} \hspace{15mm} \includegraphics[height=0.14\textheight]{../talk-oxford/images/coarse-grid.png} \,$\Omega^H=$ coarse mesh
\end{center}

\begin{align*}
&\text{pre-smooth over fine:} & & \text{smoother on } w^h \\
&\text{restrict:}                   & &\ell^H = F^H(\iR w^h) + R\, r^h(w^h) \\
&\text{coarse solve:}               & &F^H(w^H) = \ell^H \\
&\text{correct:}                    & &w^h \leftarrow w^h + P(w^H - \iR w^h) \\
&\text{post-smooth over fine:} & & \text{smoother on } w^h
\end{align*}

\bigskip
{\small
\begin{itemize}
\item $R: (\mathcal{V}^h)' \to (\mathcal{V}^H)'$ is \emph{canonical restriction}
\item $P: \mathcal{V}^H \to \mathcal{V}^h$ is \emph{canonical prolongation}
\item $\iR: \mathcal{V}^h \to \mathcal{V}^H$ is \emph{injection}
\item (restrict$+$coarse solve$+$correct) \, $=$ \, \emph{FAS coarse grid correction}
\end{itemize}
}
\end{frame}

@@

\begin{itemize}
\item<2> one can measure work in flops or run-time
\item<2> constant iterations of $O(N)$ method $\implies$ optimal
\end{itemize}
\end{frame}

@@

\only<1>{recall:

\centerline{\includegraphics[width=0.9\textwidth]{../talk-oxford/images/obstacle65.pdf}}}
\only<2>{\includegraphics[width=0.75\textwidth]{figs/ballitersV.png}}\only<3>{\includegraphics[width=0.75\textwidth]{figs/ballitersVlog.png}}\only<4>{\includegraphics[width=0.75\textwidth]{figs/ballitersVF.png}}\only<5>{\includegraphics[width=0.75\textwidth]{figs/bothitersVF.png}} \hfill
\only<2-4>{\includegraphics[width=0.2\textwidth]{../paper/fixfigs/ball-set.png}}%
\only<5>{\includegraphics[width=0.2\textwidth]{../paper/fixfigs/spiral-set.png}}

\only<2-5>{\noindent \scriptsize
\emph{all results from Bueler \& Farrell (2024)}}
\end{frame}

@@

\begin{frame}{iterations over a constraint decomposition}

\begin{itemize}
\item Tai proposed abstract iterations for solving $VI(F,\ell,\mathcal{K})$ over a CD
\item specifically the multiplicative/serial iteration:

{\small
\begin{pseudo}[left-margin=-5mm]
\pr{cd-mult}(u)\text{:} \\+
    for $i = 0,\dots,m-1$: \\+
        find $w_i\in \cK_i$ s.t. \\+
            $\displaystyle \Big<F\Big(\sum_{j<i} w_j + w_i + \sum_{j>i} \Pi_j u\Big),\, v_i - w_i\Big> \ge \ip{\ell}{v_i - w_i} \,\forall v_i \in \cK_i$ \\--
    return $w=\sum_i w_i\in\cK$
\end{pseudo}
}

\item \dots but not practical because you evaluate residuals on finest level
\item two techniques make a practical solver (Bueler \& Farrell, 2024):

    \begin{itemize}
    \item[$\circ$] \emph{defect obstacles} on each level  \hfill {\scriptsize Gr\"aser \& Kornhuber (2009)}
    \item[$\circ$] \emph{FAS coarse corrections}  \hfill {\scriptsize Brandt (1977)}
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item suppose $\mathcal{K} = \{v \ge \psi\}$ in an obstacle problem
\begin{block}{definition}
for finest-level admissible set $\mathcal{K}^J = \{v^J\ge \psi^J\} \subset \mathcal{V}^J$ and an iterate $w^J \in \mathcal{K}^J$, the \emph{defect obstacle} is
    $$\chi^J = \psi^J - w^J \in \mathcal{V}^J$$
\end{block}

    \begin{itemize}
    \item[$\circ$] note $\chi^J \le 0$
    \end{itemize}
\item generate the CD through

defect obstacles $\chi^j$ on each

level via \emph{monotone restriction}:

@@

    \begin{itemize}
    \item[$\circ$] a \emph{nonlinear} operator
    \end{itemize}
\end{itemize}

@@

\vspace{-2mm}
\begin{pseudo}[font=\small]
\pr{fascd-vcycle}(J,\ell^J,\psi^J;w^J)\text{:} \\+
    $\chi^J = \psi^J - w^J$ \\
    for $j=J$ downto $j=1$ \\+
      $\chi^{j-1} = \maxR \chi^j$ \\
      $\phi^j = \chi^j - P\chi^{j-1}$ \\
      $y^j = 0$ \\
      $\text{\pr{smooth}}^{\text{\id{down}}}(\ell^j,\phi^j,w^j;y^j)$ \\
      $w^{j-1} = \iR(w^j + y^j)$ \\
      $\ell^{j-1} = f^{j-1}(w^{j-1}) + R \left(\ell^j - f^j(w^j+y^j)\right)$ \\-
    $z^0 = 0$ \\
    $\text{\pr{solve}}(\ell^0,\chi^0,w^0;z^0)$  \\
    for $j=1$ to $j=J$ \\+
      $z^j = y^{j} + P z^{j-1}$ \\
      $\text{\pr{smooth}}^{\text{\id{up}}}(\ell^j,\chi^j,w^j;z^j)$  \\-
    return $w^J+z^J$
\end{pseudo}

\vspace{-2mm}
\tiny
\begin{itemize}
\item[] \emph{(unilateral (lower obstacle only) version shown for simplicity)}
\end{itemize}
\end{frame}

\begin{frame}{V-cycle: visualization on a 1D problem}

@@

see Bueler \& Farrell (2024) for:
\begin{itemize}
\item generalization to upper and lower obstacles:
    $$\mathcal{K}^J = \{\underline{\psi}^J \le v^J \le \overline{\psi}^J\}$$
\item stopping criteria
    \begin{itemize}
    \item[$\circ$] evaluates CP/KKT conditions
    \end{itemize}
\item FMG cycle
\item details of the $O(m_j)$ smoother used in the results: a few iterations of projected, reduced-space Newton with a few iterations of CG+ICC or GMRES+ILU on the linear step equations
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item FASCD = new multilevel solver for VI (free-boundary) problems
    \begin{itemize}
    \item[$\circ$] implemented in Python using the Firedrake library
    \item[$\circ$] observed optimality, and actually fast on tested problems
    \end{itemize}
\item many things \textbf{{\color{FireBrick} to do}}:
    \begin{itemize}
    \item[$\circ$] implement in C within PETSc
    \item[$\circ$] prove convergence; compare (Reusken, 1988)
    \item[$\circ$] mesh adaptivity, better smoothers, parabolic examples, \dots
    \end{itemize}
\end{itemize}

\bigskip\bigskip
\begin{minipage}[T]{0.25\textwidth}
\vspace{0pt}

\includegraphics[width=0.8\textwidth]{figs/QRbuelerfarrell.png}

\scriptsize paper in

SIAM J.~Sci.~Comput.
\end{minipage}
\hfill
\begin{minipage}[T]{0.25\textwidth}
\vspace{0pt}

@@

\scriptsize \hfill code at bitbucket

\phantom{foo}
\end{minipage}
\end{frame}

>>>>

\setbeamertemplate{page number in head/foot}[appendixframenumber]

@@

	\begin{itemize}
	\item[$\circ$] $\mathcal{V}=H^1(\Omega)$, $F : \mathcal{V} \to \mathcal{V}'$ continuous bijection, $\ell\in \mathcal{V}'$
	\end{itemize}
\item finite element (FE) discretization gives algebraic system on fine mesh $\Omega^h$:
    $${\color{FireBrick} F^h(u^h) = \ell^h}$$
\item suppose $w^h$ is a not-yet-converged iterate:
    $$r^h=\ell^h - F^h(w^h), \qquad \|r^h\| > \text{TOL}$$
\end{itemize}
}
\only<2>{
\begin{itemize}
\item how can we improve $w^h$ \emph{without} globally linearizing $F^h$?

	\begin{itemize}
	\item alternatives to Newton's method?
	\end{itemize}
\item rewrite the equation as the nonlinear \emph{correction equation}
    $${\color{FireBrick} F^h(u^h) - F^h(w^h) = r^h}$$
\item \underline{for $F^h$ linear}, convert the correction eqn to the \emph{error equation}
    $$F^h(e^h) = r^h,$$
with approximate solution $\tilde e^h$, and update $w^h \leftarrow w^h+\tilde e^h$
\end{itemize}

\vspace{6mm}
}
\only<3>{
\begin{itemize}
\item there are fast \emph{smoothers}: inexpensive algorithms which improve $w^h$ by efficiently removing high-frequency error components
	\begin{itemize}
	\item nonlinear Gauss-Seidel: sweep through mesh and solving nodewise problems with e.g.~1-variable Newton
	\end{itemize}
\item but how about low-frequency error components?
\item \emph{idea}: use a coarser mesh $\Omega^H$ to estimate $u^h$ in the correction eqn
\end{itemize}

\vspace{26.5mm}
}
\only<4>{
\begin{itemize}
\item Brandt's (1977) \emph{full approximation scheme} (FAS) equation:
	$${\color{FireBrick} F^H(u^H) - F^H(\iR w^h) = R \, r^h(w^h)}$$

    \begin{itemize}
    \item[$\circ$] $\iR:\mathcal{V}^h \to \mathcal{V}^H$ is node-wise \emph{injection}
    \item[$\circ$] $R:(\mathcal{V}^h)' \to (\mathcal{V}^H)'$ is \emph{canonical restriction}
    \item[$\circ$] if $w^h=u^h$ exactly then $u^H = \iR w^h$ since $F^H$ injective
    \end{itemize}

\item let $\ell^H = F^H(\iR w^h) + R\, r^h(w^h)$
\item rewritten FAS equation:
    $${\color{FireBrick} F^H(u^H) = \ell^H}$$
\end{itemize}

\vspace{2mm}
}
\end{frame}

@@

\begin{center}
fine mesh $\Omega^h$ \, \includegraphics[height=0.14\textheight]{../talk-oxford/images/fine-grid.png} \hspace{15mm} \includegraphics[height=0.14\textheight]{../talk-oxford/images/coarse-grid.png} \,coarse mesh $\Omega^H$
\end{center}

\begin{align*}
&\text{1. pre-smooth:}   & & \text{smoother on } w^h \\
&\text{2. restrict:}     & &\ell^H = F^H(\iR w^h) + R\, r^h(w^h) \\
&\text{3. coarse solve:} & & \text{solve } F^H(w^H) = \ell^H \\
&\text{4. correct:}      & &w^h \leftarrow w^h + P(w^H - \iR w^h) \\
&\text{5. post-smooth:}  & & \text{smoother on } w^h
\end{align*}

\bigskip
{\small
\begin{itemize}
\item $R: (\mathcal{V}^h)' \to (\mathcal{V}^H)'$ is \emph{canonical restriction}
\item $P: \mathcal{V}^H \to \mathcal{V}^h$ is \emph{canonical prolongation}
\item $\iR: \mathcal{V}^h \to \mathcal{V}^H$ is \emph{injection}
\item 2,3,4 \, $=$ \, \emph{FAS coarse grid correction}
\end{itemize}
}
\end{frame}

@@

\begin{itemize}
\item<2> one can measure work in flops or run-time
\item<2> constant $O(1)$ iterations of $O(N)$ method $\implies$ optimal
\end{itemize}
\end{frame}

@@

\only<1>{\includegraphics[width=0.75\textwidth]{figs/ballitersV.png}}\only<2>{\includegraphics[width=0.75\textwidth]{figs/ballitersVlog.png}}\only<3>{\includegraphics[width=0.75\textwidth]{figs/ballitersVF.png}}\only<4>{\includegraphics[width=0.75\textwidth]{figs/bothitersVF.png}} \hfill
\only<1-3>{\includegraphics[width=0.2\textwidth]{../paper/fixfigs/ball-set.png}}%
\only<4>{\includegraphics[width=0.2\textwidth]{../paper/fixfigs/spiral-set.png}}

\only<1-4>{\noindent \scriptsize \emph{all results from Bueler \& Farrell (2024)}}
\end{frame}

@@

\begin{frame}{iterations over a multilevel constraint decomposition}

\begin{itemize}
\item Tai (2003) proposed abstract iterations for solving $VI(F,\ell,\mathcal{K})$ over a CD
\item specifically the multiplicative/serial iteration:
\begin{pseudo}[left-margin=-3mm,font=\small]
\pr{cd-mult}(u)\text{:} \\+
    for $i = 0,\dots,m-1$: \\+
        find $w_i\in \cK_i$ s.t.~$\forall v_i \in \cK_i$ \\+
            $\displaystyle \Big<F\Big(\sum_{j<i} w_j + w_i + \sum_{j>i} \Pi_j u\Big),\, v_i - w_i\Big> \ge \ip{\ell}{v_i - w_i}$ \\--
    return $w=\sum_i w_i\in\cK$
\end{pseudo}

\item \dots but not practical for a multilevel CD because you evaluate residuals on finest level
\item two innovations make a practical solver (Bueler \& Farrell, 2024):

    \begin{itemize}
    \item[$\circ$] defect obstacles on each level  \hfill {\scriptsize Gr\"aser \& Kornhuber (2009)}
    \item[$\circ$] FAS coarse corrections  \hfill {\scriptsize Brandt (1977)}
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item suppose $\mathcal{K} = \{v \ge \psi\}$, a unilateral obstacle problem
\begin{block}{definition}
given a finest-level admissible set $\mathcal{K}^J = \{v^J\ge \psi^J\} \subset \mathcal{V}^J$, and an admissible iterate $w^J \in \mathcal{K}^J$, the \emph{defect obstacle} is
    $$\chi^J = \psi^J - w^J \in \mathcal{V}^J$$
\end{block}

    \begin{itemize}
    \item[$\circ$] note $\chi^J \le 0$
    \end{itemize}
\item multilevel: generate defect

obstacles $\chi^j$ on each level

 via \emph{monotone restriction}:

@@

    \begin{itemize}
    \item[$\circ$] $R^{\oplus}$ is nonlinear
    \end{itemize}
\end{itemize}

@@

\vspace{-2mm}
\begin{pseudo}[font=\small]
\pr{fascd-vcycle}(J,\ell^J,\psi^J;w^J)\text{:} \\+
    $\chi^J = \psi^J - w^J$ \\
    for $j=J$ downto $j=1$: \\+
      $\chi^{j-1} = \maxR \chi^j$ \\
      $\phi^j = \chi^j - P\chi^{j-1}$ \\
      $y^j = 0$ \\
      $\text{\pr{smooth}}^{\text{\id{down}}}(\ell^j,\phi^j,w^j;y^j)$ \\
      $w^{j-1} = \iR(w^j + y^j)$ \\
      $\ell^{j-1} = f^{j-1}(w^{j-1}) + R \left(\ell^j - f^j(w^j+y^j)\right)$ \\-
    $z^0 = 0$ \\
    $\text{\pr{solve}}(\ell^0,\chi^0,w^0;z^0)$  \\
    for $j=1$ to $j=J$: \\+
      $z^j = y^{j} + P z^{j-1}$ \\
      $\text{\pr{smooth}}^{\text{\id{up}}}(\ell^j,\chi^j,w^j;z^j)$  \\-
    return $w^J+z^J$
\end{pseudo}

\vspace{-2mm}
\tiny \hfill \emph{unilateral (lower obstacle only) version shown for simplicity}
\end{frame}

\begin{frame}{V-cycle visualization (1D unilateral obstacle problem)}

@@

see Bueler \& Farrell (2024) for:
\begin{itemize}
\item generalization to upper and lower obstacles:
    $$\mathcal{K}^J = \{\underline{\psi}^J \le v^J \le \overline{\psi}^J\}$$
\item stopping criteria
    \begin{itemize}
    \item[$\circ$] evaluates CP/KKT conditions
    \end{itemize}
\item FMG cycle
\item details of the $O(m_j)$ smoother used in the results: a few iterations of a reduced-space Newton method, with a few iterations of CG+ICC or GMRES+ILU on the linear step equations
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item FASCD = new multilevel solver for VI problems
    \begin{itemize}
    \item[$\circ$] implemented in Python using the \href{https://www.firedrakeproject.org/}{Firedrake FE library (\texttt{firedrakeproject.org})}
    \item[$\circ$] observed optimality, and actually fast on tested problems
    \end{itemize}
{\small
\item much work to do:
    \begin{itemize}
    \item[$\circ$] implement in C within \href{https://petsc.org/release/}{PETSc (\texttt{petsc.org})}
    \item[$\circ$] prove convergence; compare (Reusken, 1988)
    \item[$\circ$] mesh adaptivity, better smoothers, parabolic examples, \dots
    \end{itemize}
}
\end{itemize}

\bigskip\bigskip
\footnotesize

\noindent \href{https://epubs.siam.org/doi/10.1137/23M1594200}{\texttt{doi/10.1137/23M1594200}} \hfill \href{https://bitbucket.org/pefarrell/fascd/}{\texttt{bitbucket.org/pefarrell/fascd/}}

\smallskip
\begin{minipage}[T]{0.35\textwidth}
\vspace{0pt}

\includegraphics[width=0.57\textwidth]{figs/QRbuelerfarrell.png}

\scriptsize the paper (SIAM J.~Sci.~Comput.)
\end{minipage}
\hfill
\begin{minipage}[T]{0.25\textwidth}
\vspace{0pt}

@@

\scriptsize \hfill the code
\end{minipage}
\end{frame}

\appendix
